---
title: "EuroBioC2025 OSTA Workshop"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{EuroBioC2025 OSTA Workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE, 
  message=FALSE
)
```

# Workflow: Visium HD cell-level

Authors: Yixing E. Dong[^1], Ellis Patrick[^2].

[^1]: University of Lausanne, Lausanne, Switzerland

[^2]: University of Sydney, Sydney, Australia

Last modified: 20 January, 2026.

## Preparation

### Dataset

The relevant data by @oliveira2025high are stored on Open Science Framework, and can be downloaded as delayed format cached on your laptop through `r BiocStyle::Biocpkg("OSTA.data")`:

```{r load-data-test, message=FALSE}
library(OSTA.data)
id <- "VisiumHD_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
list.files(file.path(td, "segmented_outputs"))
```

We prepared a reader for Visium HD segmented output at `./vignettes/utils.R`. 

**Update (January 2026):** Pull requests branched from this reader have been merged into the `VisiumIO` and `SpatialFeatureExperiment` packages.

### *R* / *Bioconductor* packages used

Bioconductor: Biobase, BiocStyle, Banksy, CARDspa, DropletUtils, GenomeInfoDb, ggspavis, OSTA.data, S4Vectors, scater, scran, scuttle, SingleCellExperiment, spacexr, SpatialExperiment, SpatialFeatureExperiment, spicyR, SpotSweeper, Statial, Voyager

CRAN: dplyr, ggplot2, grDevices, grid, here, igraph, magick, pals, patchwork, pheatmap, plotly, rlang, sf, tidyr, tidySpatialExperiment, tidyverse

## Workshop goals and objectives

The key steps of Visium HD data analysis are described in this workshop, along with diverse visualizations of the methods used. The goal is to help beginner bioinformaticians working with Visium HD data become familiar with each step involved and complete a standard analysis pipeline from start to finish.

### Learning goals

-   Learn how to analyze Visium HD data with cell segmentation output
-   Understand the key steps and methods for deriving biological insights
-   Create aesthetic visualizations of Visium HD data using existing R packages

### Learning objectives

-   Read Visium HD segmentation output data as a `SpatialFeatureExperiment` object
-   Understand spatially aware quality control metrics
-   Perform label transfer from single-cell reference data to spatial data
-   Identify spatial domains and marker genes for each region
-   Visualize Visium HD data as polygons or cell centroids
-   Extract morphological metrics from cell polygons and make inference
-   Conduct spatial analysis in a tissue slice

### Time outline

| Activity              | Time    |
|-----------------------|---------|
| Introduction          | 5 mins  |
| Initialization        | 3 mins  |
| Data loading          | 7 mins  |
| Data processing       | 5 mins  |
| Quality control       | 10 mins |
| Cell annotation       | 10 mins |
| Shape statistics      | 15 mins |
| Spatial analysis      | 25 mins |
| Conclusion            | 5 mins  |

## Introduction

As of June 2025, Visium HD enables direct output of H&E-segmented cell-level data from [SpaceRanger v4](https://www.10xgenomics.com/analysis-guides/segmentation-visium-hd). This automation removes the need for an additional segmentation tool, such as `bin2cell`. Since biologists often find cellular units more intuitive for deriving insights, this output may attract more interest than bin-level data. However, bins outside cell boundaries can still provide valuable information for many research topics in the tumor microenvironment. In this workflow, we will demonstrate several cell-level analysis steps using the same dataset, subsetted to the same region, as in the *Sequencing-based platforms - Workflow: Visium HD bin-level* chapter of the OSTA book.

### Dependencies {#sec-seq-workflow-visium-hd-cell-load-data}

```{r deps, message=FALSE, warning=FALSE}
library(AUCell)
library(Biobase)
library(BiocStyle)
library(Banksy)
library(CARDspa)
library(dplyr)
library(DropletUtils)
library(GenomeInfoDb)
library(ggfortify)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(grDevices)
library(grid)
library(here)
library(igraph)
library(magick)
library(msigdbr)
library(OSTA.data)
library(pals)
library(patchwork)
library(pheatmap)
library(plotly)
library(rlang)
library(S4Vectors)
library(scater)
library(scran)
library(scuttle)
library(sf)
library(SingleCellExperiment)
library(sosta)
library(spacexr)
library(SpatialExperiment)
library(spatialFDA)
library(SpatialFeatureExperiment)
library(spatstat)
library(spicyR)
library(SpotSweeper)
library(Statial)
library(stats)
library(tidyr)
library(tidySpatialExperiment)
library(tidyverse)
library(Voyager)
```

We source the reader defined in `utils.R` and specify alias for plotting themes. 

```{r prep}
source(here::here("vignettes/utils.R"))

# ggplot theme alias
center_title <- theme(plot.title = element_text(hjust = 0.5))

no_legend <- theme(legend.position = "none")

gradient_theme <- theme(plot.title = element_text(hjust = 0.5),
                        legend.title = element_blank(),
                        legend.key.width = grid::unit(0.5, "lines"), 
                        legend.key.height = grid::unit(1, "lines"))

thin_discrete_legend <- theme(legend.key.size = grid::unit(0.5, "lines"))

enlarge_legend_dot <- guides(col=guide_legend(override.aes=list(size=3)))

common_args_heatmap <- list(cellwidth=10, cellheight=10, 
                            treeheight_row=5, treeheight_col=5)
```

### Load spatial data

Output files from SpaceRanger v4 are located in `/segmented_outputs` folder, following the expected structure described in the *Background - Importing data* chapter of the OSTA book.

```{r osta_download}
id <- "VisiumHD_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
```

<!-- Behind the scenes, we begin by reading the cell segmentation as an `sf` object, along with the scaling factor `.json` file and the image `.png` file. We use the high resolution image, as it provides the expected H&E detail beneath the segmented cell masks. The filtered count matrix (`.h5`) is read in as a `SingleCellExperiment`. We then construct a `SpatialExperiment` object using the components above, where cell centroids are derived from the `sf` object and used as spatial coordinates. Next, we coerce the `SpatialExperiment` into a `SpatialFeatureExperiment`, with the `sf` polygons stored in `colGeometries()`. To align the polygons with the orientation of the H&E image, we flip the y-axis of both the geometreis and the spatial coordinates. Finally, we store the scaled coordinates in `colData()` to facilitate easier subsetting. 
-->

We read the Visium HD cell-segmented data into a `SpatialFeatureExperiment` object.

```{r load-other-vhdcell, message=FALSE, warning=FALSE}
vhdsfe <- readVisiumHDCellSeg(td, res = "hires")
vhdsfe
```

To keep the runtime low, downstream analyses are performed on a subset region corresponding to 16 µm bins, representing approximately 1/64 of the full data coverage area. We restrict the analysis to the same region used in the OSTA book.

```{r subset-polygons, message=FALSE, warning=FALSE}
vhdsub <- subsetVisiumHD(vhdsfe,
                         xmin = 4214.937, xmax = 4465.761,
                         ymin = 2979.286, ymax = 3211.817)
```

Because the analysis is limited to this subset of the tissue, the number of cells in the corresponding `sf` object is expected to be reduced by roughly the same fraction, depending on local tissue density.

```{r dim-polygons, message=FALSE, warning=FALSE}
ncol(vhdsfe)/ncol(vhdsub)
```

Show the subset region used for analysis:

```{r plt-subset-cov-box, fig.width=6, fig.height=4, message=FALSE}
plotImage(vhdsfe) + 
  plotBox(getBoxRange(vhdsfe), col = "#3b3b3b", lty = 2, lwd = 1/2) +
  plotBox(getBoxRange(vhdsub), col = "black", lty = 4, lwd = 2/3) +
  ggtitle("Grey: data coverage; Black: subset region") + 
  center_title
```

Plot cell polygons overlaid on the H&E image for this subset region (in black box). Here is an overview of the tissue subset that we will use throughout the workflow.

```{r plt-subset-he-polygon, fig.width=8, fig.height=4, message=FALSE}
vhdsub$polygon <- 1
p0 <- plotImage(vhdsub, bbox = unlist(getBoxRange(vhdsub))) 

p1 <- plotSpatialFeature(vhdsub, bbox = unlist(getBoxRange(vhdsub)),
                         features = "polygon", alpha = 0.1, fill = NA, color = "white",
                         colGeometryName = "updatecellSeg", image_id = "hires") + 
  no_legend

(p0 + ggtitle("H&E region") |
  p1 + ggtitle("Overlayed with cell boundaries")) &
  center_title
```

```{r hidden1, echo = FALSE, results = "hide"}
gc()
```

## Which cells pass quality check?

Compute cell-level QC metrics using `r BiocStyle::Biocpkg("scuttle")`.

```{r qc-metrics-vhdsub, message=FALSE, warning=FALSE}
# Visualize library size to the polygons
sub <- list(mt = grep("^MT-", rownames(vhdsub)))
vhdsub <- addPerCellQCMetrics(vhdsub, subsets = sub)
vhdsub$log_sum <- log1p(vhdsub$sum)

CD <- colData(vhdsub) %>% as.data.frame()
head(CD %>% select(Barcode, sum, log_sum, subsets_mt_percent))
```

Here is an overview of cells colored by their log library size and mitochondrial percentage.

```{r loglib-mitopct, fig.width=8, fig.height=4, message=FALSE}
p0 <- plotSpatialFeature(vhdsub, features = "log_sum", 
                   colGeometryName = "updatecellSeg") + 
  ggtitle("log library size") 

p1 <- plotSpatialFeature(vhdsub, features = "subsets_mt_percent", 
                         colGeometryName = "updatecellSeg") + 
  ggtitle("% mitochondrial") 

(p0 | p1) &
  gradient_theme & 
  scale_fill_gradientn(colors = pals::jet())
```

### SpotSweeper

`r BiocStyle::Biocpkg("SpotSweeper")` identifies outliers based on low log-library size, a low number of uniquely detected features, or a high mitochondrial fraction relative to the surrounding neighborhood:

```{r qc-vhdsub-local}
vhdsub <- localOutliers(vhdsub, metric="sum", direction="lower", log=TRUE)
vhdsub <- localOutliers(vhdsub, metric="detected", direction="lower", log=TRUE)
vhdsub <- localOutliers(vhdsub, metric="subsets_mt_percent", direction="higher", log=TRUE)
vhdsub$discard <- 
    vhdsub$sum_outliers | 
    vhdsub$detected_outliers | 
    vhdsub$subsets_mt_percent_outliers
# tabulate number of cells retained vs. removed by any criterion 
table(vhdsub$discard)
```

We can visualize the local outliers identified by `SpotSweeper` in a spatial context. However, they are difficult to discern, as only a small number of relatively small cells are flagged as outliers.

```{r qc-vhdsub-plt, fig.width=8, fig.height=3, message=FALSE}
common_args_qc <- list(fill = NA, color = "grey90", linewidth = 0.1, 
                    colGeometryName = "updatecellSeg")

p0 <- exec(plotSpatialFeature, 
           vhdsub, features = "sum_outliers", !!!common_args_qc) +
  ggtitle("low_lib_size") 

p1 <- exec(plotSpatialFeature, 
           vhdsub, features = "detected_outliers", !!!common_args_qc) +
  ggtitle("low_n_features") 

p2 <- exec(plotSpatialFeature, 
           vhdsub, features = "discard", !!!common_args_qc) +
  ggtitle("discard")

p0 + p1 + p2 +
  plot_layout(nrow=1, guides="collect") &
  thin_discrete_legend & center_title &
  scale_fill_manual("discard", values=c("white", "red"))

```

```{r hidden2, echo = FALSE, results = "hide"}
gc()
```

## What type of cells do we have in this region?

### Load reference

This Visium HD dataset includes matched scRNA-seq data from the same tissue, which helps avoid tissue-specific and batch effects that often complicate label transfer. The `Level1` column contains broad, interpretable cell populations that are well suited for deconvolution and downstream interpretation. We use these labels to map likely cell identities onto the Visium HD cells.

First, we load the data from the `OSTA.data` package.

```{r ann-load-sce, message=FALSE, warning=FALSE}
## SCE ref
id <- "Chromium_HumanColon_Oliveira"

## Download the data for OSTA.data and unzip the files
pa <- OSTA.data_load(id)
dir.create(td2 <- tempfile())
unzip(pa, exdir=td2)

# read into 'SingleCellExperiment'
fs <- list.files(td2, full.names=TRUE)
h5 <- grep("h5$", fs, value=TRUE)
sce <- read10xCounts(h5, col.names=TRUE)

# add cell metadata
csv <- grep("csv$", fs, value=TRUE)
cd <- read.csv(csv, row.names=1)
colData(sce)[names(cd)] <- cd[colnames(sce), ]

# use gene symbols as feature names
rownames(sce) <- make.unique(rowData(sce)$Symbol)

# exclude cells deemed to be of low-quality
sce <- sce[, sce$QCFilter == "Keep"]
```

We will use this single-cell dataset as a reference. It contains cell annotations at two resolutions. First, let’s inspect the labels in `Level1.`

```{r ann-level1-table}
# tabulate subpopulations
table(sce$Level1)
```

Spaces can make manipulating data tricky sometimes, so let's replace them with underscores.

```{r ann-sanitize-level1}
sce$Level1 <- gsub(" ", "_", sce$Level1)
```

For speed, we down-sample to at most 1,000 cells per reference class. This preserves diversity within each class while keeping memory use and runtime manageable.

```{r ann-downsample-sce, message=FALSE, warning=FALSE}
# (this is done only to keep runtime/memory low)
cells_by_type <- split(seq_len(ncol(sce)), sce$Level1)
down_sample <- unlist(lapply(cells_by_type, \(.) sample(., min(length(.), 1000))))
```

### RCTD - Robust decomposition of cell type mixtures in spatial transcriptomics

There are many approaches for transferring labels from single-cell data to spatial data. Here, we use `r BiocStyle::Biocpkg("spacexr")` (i.e., `RCTD`) because it performs competitively in public benchmarks (@sang2024spotless; @li2023comprehensive; @gaspard2025cell), explicitly models mixed signals, and assigns each spatial unit a clear classification as either a singlet or one of several doublet types.

`RCTD` was originally developed for spot-based platforms such as 10x Visium, where each spot captures RNA from multiple cells. The method models the observed counts in each spot as a mixture of reference cell-type expression profiles and uses a likelihood-based framework to estimate non-negative weights for each candidate cell type. It then compares a singlet model with a doublet model to determine whether the counts are best explained by one or two cell types. In addition, `RCTD` learns scale factors to account for differences in platform characteristics and total RNA content, facilitating robust label transfer across technologies (@cable2022robust). 

In our setting, we apply `RCTD` at cellular resolution using segmented Visium HD data. Although each unit is intended to represent a single cell, there are cases where the measured counts may reflect contributions from multiple cell types. This can arise from partial-volume effects at cell boundaries, segmentation errors, densely packed regions, or local diffusion of transcripts. Running `RCTD` in doublet mode remains useful in this context, as it can flag such cases and provide a principled estimate of the most likely composition. In practice, we expect most units to be classified as singlets, with a smaller fraction of doublets that warrant closer inspection or additional filtering.

Because running `RCTD` is computationally intensive, we load a precomputed result for this workflow.

```{r rctd-preload-level1}
# data("rctd_result_level1.RData", package = "EuroBioC2025_OSTA_Workshop")
load(here::here("data/rctd_result_level1.RData"))
```

!!! Do not run this in the workshop, as it will take 30 mins.

```{r rctd-reproduce-run, warning=FALSE, message=FALSE, eval=FALSE}
# ## Perform initial preprocessing of the visiumHD and single cell reference data
# rctd_data <- createRctd(spatial_experiment = vhdsub,              # Our visium HD data
#                         reference_experiment = sce[, down_sample],# Down-sampled single cell reference
#                         cell_type_col="Level1",                   # We will use the Level1 resolution of cell types.
#                         UMI_min = 0)                              # To make things easier, we won't filter away cells that RCTD thinks are poor quality.
# 
# ## Perform the deconvolution.
# rctd_result_level1 <- runRctd(rctd_data, max_cores = 4, rctd_mode = "doublet")
# # save(rctd_result_level1, file = here::here("data/rctd_result_level1.RData"))
```

Note: `RCTD` includes internal quality control steps, so we first check whether any cells were filtered out before running the deconvolution.

```{r rctd-qc-dims}
dim(vhdsub)
dim(rctd_result_level1)
setdiff(colnames(vhdsub), colnames(rctd_result_level1))
vhdsub$discard[colnames(vhdsub) == "91874"]
vhdsub$sum_outliers[colnames(vhdsub) == "91874"]
vhdsub$sum[colnames(vhdsub) == "91874"]
head(sort(vhdsub$sum))
```

The spot labels are stored in `colData()` as `spot_class`. This provides a quick quality check: observing a majority of singlet classifications at cellular resolution is expected and supports the validity of the H&E-guided segmentation.

```{r rctd-spot-class-tab, warning=FALSE, message=FALSE}
table(rctd_result_level1$spot_class)
```

We then synchronize the deconvolution labels back into our `vhdsub` `SpatialFeatureExperiment` object by matching cell IDs, ensuring that all components remain properly aligned.

```{r rctd-assign-celltype}
# Subset to RCTD filtered object for convenience
vhdsub <- vhdsub[, colnames(vhdsub)]

# Here we use match to ensure that we match the cell IDs
vhdsub$cellTypeClass <- rctd_result_level1$spot_class[
  match(colnames(vhdsub),colnames(rctd_result_level1))]
vhdsub$cellType <- rctd_result_level1$first_type[
  match(colnames(vhdsub), colnames(rctd_result_level1))]

# Double check that the numbers of the cell types is consistent
table(vhdsub$cellType)
```

We can visualize the spatial distribution of singlets, doublets, or the cell types assigned by `RCTD`.

```{r rctd-result-discrete, fig.width=8, fig.height=3, message=FALSE}
p0 <- plotSpatialFeature(vhdsub, features = "cellTypeClass", 
                   colGeometryName = "updatecellSeg") + 
  scale_fill_manual(values = unname(pals::okabe()))
p1 <- plotSpatialFeature(vhdsub, features = "cellType", 
                   colGeometryName = "updatecellSeg") + 
  scale_fill_manual(values = unname(pals::trubetskoy()))

(p0 | p1) & thin_discrete_legend
```

Previously, we identified a cell that was flagged as a local outlier by `SpotSweeper`. We can now examine its properties in more detail.

```{r rctd-agree-spotsweeper}
table(vhdsub$subsets_mt_percent_outliers == TRUE)
vhdsub$cellType[vhdsub$subsets_mt_percent_outliers == TRUE]
vhdsub$cellTypeClass[vhdsub$subsets_mt_percent_outliers == TRUE]
```

Filter out low-quality cells from the `SpatialFeatureExperiment` object prior to performing spatial clustering.

```{r qc-filtering, message=FALSE, warning=FALSE}
vhdsub <- vhdsub[, !vhdsub$discard]
dim(vhdsub)
```

```{r hidden3, echo = FALSE, results = "hide"}
gc()
```

## Spatial domain identification

First, we identify highly variable genes (HVGs) within the target region. Note that spatially variable genes (SVGs) could be used as an alternative.

```{r vhdsub-hvg}
vhdsub <- logNormCounts(vhdsub)
dec <- modelGeneVar(vhdsub)
hvg <- getTopHVGs(dec, n=3e3)
```

### Banksy

`r BiocStyle::Biocpkg("Banksy")` utilizes a pair of spatial kernels to capture gene expression variation, followed by dimension reduction and graph-based clustering to identify spatial domains.

```{r Banksy, message=FALSE}
# restrict to selected features
tmp <- vhdsub[hvg, ]

# compute spatially aware 'Banksy' PCs
set.seed(1000)
tmp <- computeBanksy(tmp, 
                     assay_name = "logcounts", 
                     coord_names = c("array_row", "array_col"),  
                     k_geom = 8)   # consider first order neighbors
tmp <- runBanksyPCA(tmp, 
                    lambda = 0.2, # use little spatial information, 
                    npcs = 20)
reducedDim(vhdsub, "PCA") <- reducedDim(tmp)

set.seed(1000)
tmp <- Banksy::clusterBanksy(tmp, lambda = 0.2, npcs = 20, resolution = 0.8)
vhdsub$Banksy <- tmp$clust_M0_lam0.2_k50_res0.8

table(vhdsub$Banksy)
```

We can now visualize the clustering results spatially.

```{r clus-spa-umap, fig.width=3.5, fig.height=3, warning=FALSE, message=FALSE}
plotSpatialFeature(vhdsub, features = "Banksy", 
                   colGeometryName = "updatecellSeg") +
  scale_fill_manual(values = unname(pals::kelly())) + 
  thin_discrete_legend
```

Next, we perform differential gene expression (DGE) analysis to identify marker genes for each cluster. We also compute the cluster-wise mean expression of selected markers.

```{r mgs}
# differential gene expression analysis
mgs <- findMarkers(vhdsub, groups = vhdsub$Banksy, direction = "up")
# select for a few markers per cluster
top <- lapply(mgs, \(df) rownames(df)[df$Top <= 2])
top <- unique(unlist(top))
# average expression by clusters
pbs <- aggregateAcrossCells(vhdsub, 
    ids = vhdsub$Banksy, subset.row = top, 
    use.assay.type = "logcounts", statistics = "mean")
```

Marker genes for each cluster can be visualized using a heatmap.:

```{r plt-mgs-hm, message=FALSE, fig.width=4.5, fig.height=2.5}
# visualize averages z-scaled across clusters
exec(pheatmap,
    mat = t(assay(pbs)), scale = "column", breaks = seq(-2, 2, length = 101),
    !!!common_args_heatmap)
```

Here, we highlight selected marker genes from clusters 1, 3, and 6, and also visualize their expression patterns in a spatial context:

```{r plt-mgs-xy, fig.width=8, fig.height=2.5, message=FALSE}
p0 <- plotSpatialFeature(vhdsub, features = "MMP2", 
                   colGeometryName = "updatecellSeg") + ggtitle("MMP2") 

p1 <- plotSpatialFeature(vhdsub, features = "PIGR", 
                         colGeometryName = "updatecellSeg") + ggtitle("PIGR") 

p2 <- plotSpatialFeature(vhdsub, features = "DES", 
                         colGeometryName = "updatecellSeg") + ggtitle("DES") 

(p0 | p1 | p2) &
  gradient_theme & 
  scale_fill_gradientn(colors = rev(hcl.colors(9, "Rocket")))

```

```{r hidden4, echo = FALSE, results = "hide"}
gc()
```

## Concordance of deconvolution & clustering

Here, we juxtapose the spatial plots of the deconvolution and clustering results.

```{r plt-dec-clu-xy, fig.width=8, fig.height=3, message=FALSE}
p0 <- plotSpatialFeature(vhdsub, features = "cellType", 
                         colGeometryName = "updatecellSeg") + 
  scale_fill_manual(values = unname(pals::trubetskoy())) + 
  ggtitle("RCTD") 

p1 <- plotSpatialFeature(vhdsub, features = "Banksy", 
                   colGeometryName = "updatecellSeg") + 
  scale_fill_manual(values = unname(pals::kelly())) + 
  ggtitle("Banksy")

p0 + p1 +
  plot_layout(nrow=1) &
  center_title & thin_discrete_legend
```

We observe that `Banksy` delineates tissue boundaries effectively, while `RCTD` provides direct biological insight into the underlying clusters. The concordance between the two methods can be visualized using a heatmap:

```{r plt-dec-clu-hm, fig.width=2.5, fig.height=3}
#| code-fold: true
fq <- prop.table(table(vhdsub$Banksy, vhdsub$cellType), 1)
exec(pheatmap, fq, !!!common_args_heatmap)
```

```{r hidden5, echo = FALSE, results = "hide"}
gc()
```

## Feature-set signatures

Individual biomarkers identified in the DGE step may or may not have direct phenotypic relevance. Instead, we can quantify gene sets that are known to orchestrate biological functions or pathways, such as metabolic activity, cell cycle progression, and cell death, at the single-cell level.

MSigDB provides a programmatic interface to the [Molecular Signatures Database (MSigDB)](https://www.gsea-msigdb.org/gsea/msigdb) (@subramanian2005gene), one of the largest curated collections of molecular signatures and pathways. We begin by retrieving the human Hallmark gene set collection.

```{r featset}
# retrieve & simplify hallmark gene sets from 'MSigDB'
db <- msigdbr(species="Homo sapiens", collection="H")
gs <- split(db$ensembl_gene, db$gs_name)
names(gs) <- tolower(gsub("HALLMARK_", "", names(gs)))
```

Next, we score these gene sets using `r BiocStyle::Biocpkg("AUCell")` (@aibar2017scenic), a rank-based approach that (i) ranks genes for each cell and (ii) computes the area under the recovery curve (AUC) for each gene set.

```{r featset-gene-ranking}
# realize (sparse) gene expression matrix
mtx <- as(logcounts(vhdsub), "dgCMatrix") 
# use ensembl identifiers as rownames
rownames(mtx) <- rowData(vhdsub)$ID
# filter for genes represented in panel
.gs <- lapply(gs, intersect, rownames(mtx))
# keep only those with at least 5 genes
.gs <- .gs[sapply(.gs, length) >= 5]
# build per-cell gene rankings
rnk <- AUCell_buildRankings(mtx, plotStats=FALSE, verbose=FALSE)
# calculate AUC for each gene set in each cell
auc <- AUCell_calcAUC(geneSets=.gs, rankings=rnk, verbose=FALSE)
```

A heatmap can be used to examine which signatures are highly expressed in each cell type. To diversify the visualization, we instead summarize the top two signatures per cell type and display their scaled average enrichment scores in a waterfall plot.

```{r sigbarplot, fig.width=7, fig.height=4.1}
# aggregate AUC values by cellType
mu <- aggregateAcrossCells(auc, ids=vhdsub$cellType,
                           use.assay.type="AUC", statistics="mean")
mu_row_scaled <- rowScale(mu)
top_mu <- getTopSigDF(mu_row_scaled, top_n = 2)

plotDivergentBar(top_mu)
```

p53 is a well-known pathway involved in tumor cell growth (@vogelstein2000surfing), and its enrichment score can be visualized spatially.

```{r p53-spatial, fig.width=3.8, fig.height=3, message=FALSE}
# add results as cell metadata
colData(vhdsub)[rownames(auc)] <- t(assay(auc)) 
plotSpatialFeature(vhdsub, feature = "p53_pathway",
                   colGeometryName = "updatecellSeg") + gradient_theme +
  ggtitle("p53 Pathway")
```

```{r hidden6, echo = FALSE, results = "hide"}
gc()
```

## Shape statistics

Because cell segmentation polygons are available for each cell, we can derive a wide range of shape metrics (@gunz2025analysis). `r BiocStyle::Biocpkg("sosta")` was originally developed to analyze multi-cellular anatomical structures using polygons that outline regions of cells; here, we extend its use to investigate the geometry of individual cells.

```{r shape-metrics}
# get cell segmentations
cellpo <- colGeometry(vhdsub, "updatecellSeg")
cellpo$cellType <- vhdsub$cellType
# get shape metrics
shme <- totalShapeMetrics(cellpo)
shme[1:3, 1:3]
```

Here we can check the relationship between shape metrics in the first two principal components (PCs):

```{r shme-pca-loading, fig.width=6,fig.height=4.2, message=FALSE,warning=FALSE}
tshme <- t(shme) |> as.data.frame() 
autoplot(
  prcomp(tshme, scale. = TRUE),
  data = cellpo,
  color = "cellType",
  size = 0.8, alpha = 0.8, 
  loadings.colour = "blue",
  loadings.label = TRUE,
  loadings.label.size = 5, 
  loadings.label.colour = "black") +
  scale_color_manual(values=unname(pals::trubetskoy())) +
  theme_classic() + enlarge_legend_dot

```

The first principal component (PC1) explains 42.82% of the variability in the shape statistics, and the second principal component (PC2) captures 31.66%. The angle between two shape metrics reflects their degree of similarity, while the length of a loading indicates the strength of a given metric. For example, the loadings of `fibreLength` and `Compactness` are nearly perpendicular, indicating a weak or negative correlation between them, which is consistent with their relationship in the scatter plot. We can also color the points by the log gene expression of *PIGR*, a common epithelial marker, as shown previously. Cells with high *PIGR* expression tend to exhibit longer `fibreLength`, with no obvious association with compactness.

```{r shape-cor, fig.width=4.5, fig.height=4}
tshme$logPIGR <- logcounts(vhdsub)["PIGR", ]
ggplot(tshme, aes(x = fibreLength, y = Compactness)) + 
  geom_point(aes(color = logPIGR), alpha = 0.8) + 
  scale_color_viridis_c(option = "rocket", direction = -1) + 
  geom_density2d() + 
  geom_smooth(method='loess', formula= y~x) + 
  stat_cor(method = "pearson", label.x.npc = "center", label.y.npc = "top") +
  theme_classic() + gradient_theme + ggtitle("PIGR")
  
```

Next, we can summarize the shape statistics per cell type. We check the differences in `fibreLength` and `Compactness` across cell types. 

```{r shape-violin, fig.width=7.2, fig.height=4.5}
tshme |>
  cbind(cellType = vhdsub$cellType) |> 
  pivot_longer(cols = -cellType, names_to = "metrics", values_to = "val") |> 
  filter(metrics %in% c("Compactness", "fibreLength")) |>
  ggplot(aes(x = cellType, y = val, fill = cellType)) +
  geom_violin() +
  geom_boxplot(aes(fill = NULL), width = 0.3) +
  facet_wrap(~metrics, scales = "free") +
  scale_fill_manual(values = unname(pals::trubetskoy())) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  guides(fill = "none")
```

We observe more uniform distribution in `Compactness`, while more variability in `fibreLength` across cell types. 

We can also view these shape metrics, such as `fibreLength`, spatially:

```{r fiblen-spatial, fig.width=3.8, fig.height=3, message=FALSE}
colData(vhdsub)[names(tshme)] <- tshme

plotSpatialFeature(vhdsub, features = "fibreLength",
                   colGeometryName = "updatecellSeg") + 
  scale_fill_gradientn(colors = c("#3B3EA8", "#278EC8", "orange", "yellow")) +
  gradient_theme + ggtitle("fibreLength")
```

For example, immune cells have shorter `fibreLength` compared to tumor and stromal cells, which is also shown in the boxplot above. 

```{r hidden7, echo = FALSE, results = "hide"}
gc()
```

## Interrogating the spatial arrangement of our cells.

What makes spatial transcriptomics more powerful than scRNA-seq is its integration with spatial statistics. Measures of spatial association are commonly classified as either global or local.

### Global spatial association

Next, we quantify spatial proximity between annotated cell types. The `getPairwise()` function from the `r BiocStyle::Biocpkg("spicyR")` package computes the K-cross function for all pairs of cell types, summarizing whether cells of type A tend to occur within a distance `r` of cells of type B more frequently than expected under spatial randomness. Here, we use a radius of 200 units (\~50 microns) to capture immediate neighborhood effects, corresponding to roughly one to two cell diameters.

```{r spicy-pairwise-kcross}
crossK = getPairwise(vhdsub, 
                     r = 200, 
                     cellType = "cellType",
                     imageID = "sample_id",
                     spatialCoords = c("pxl_col_in_fullres", 
                                       "pxl_row_in_fullres"))
```

We can visualize these relationships using `imageCrossPlot()`, which displays spatial enrichment and avoidance across all cell-type pairs. This overview is useful for scanning for strong interactions before exploring specific pairs in more detail. To aid interpretation, note the sign of the values: those above zero indicate spatial attraction at radius `r` (more co-occurrence than expected), while values below zero indicate spatial repulsion.

```{r spicy-pairwise-plot, fig.height=4, fig.width=5}
imageCrossPlot(crossK, limits = c(-400,400)) + gradient_theme +
  ggtitle("Spatial association") + 
  theme(axis.line = element_blank(), axis.ticks = element_blank())
```

We can revisit the spatial composition of cell types annotated by `RCTD` to build intuition for these statistics. This helps confirm that strong pairwise signals are visible in the tissue image and are not driven by a handful of cells.

### Local indicator of spatial association - LISA

Visium HD cell-level data can be modeled as a point pattern process (@emons2025harnessing). In contrast to the global average measures described above, local indicators of spatial association (LISA) quantify the local contribution of each cell (@anselin1995local). First, we construct a plotting data frame containing the LISA Ripley’s K values for tumor cells only.

```{r lisapltdf, message=FALSE, warning=FALSE}
df <- .speToDf(vhdsub)
pp <- .dfToppp(df, marks="cellType")

# subset to only Tumor cells 
ppSub <- subset(pp, marks %in% "Tumor")

# calculate LISA K curves
resLocal <- localK(ppSub, verbose=FALSE) 

# code adapted from 
# https://robinsonlabuzh.github.io/pasta/01-imaging-univar-ppSOD.html
df <- resLocal |>
  as.data.frame() |>
  pivot_longer(cols = starts_with("iso"), names_to="curve") 

sel <- df |>
  filter(r > 700.5630 & r < 702.4388) |>
  mutate(sel=value) |> 
  select(curve, sel)

plt_df <- left_join(df, sel)
head(plt_df)
```

Plot the LISA Ripley’s K value for each tumor cell, colored by its value at a radius of 700 pixels. Curves above the gray line that represents complete spatial randomness (CSR) indicate highly clustered tumor cells in the right spatial plot.

```{r lisa-plot, fig.height=4, fig.width=8.2, warning=FALSE, message=FALSE}
p <- ggplot(plt_df, aes(r, value, group=curve, col=sel)) +
  geom_line() +
  geom_line(aes(y=theo), linetype=2, col="darkgray") +
  geom_vline(xintercept=700) +
  scale_color_viridis_c(option = "mako") + 
  xlab("Radius (r)") + 
  scale_y_continuous(breaks = c(0, 2e6, 4e6), labels = c("0", "2", "4"),
                     name = bquote("Relationship value x"~10^6)) +
  theme_classic() + 
  theme(legend.position="none", 
        axis.title = element_text(size=14),
        panel.grid.major = element_line(colour = "grey90", linetype = 2)) 

q <- ggplot(sel %>% cbind(data.frame(x = ppSub$x, y = ppSub$y)), 
            aes(x, y, col=sel)) +
  coord_equal(expand=FALSE) +
  geom_point(size=1) + 
  scale_color_viridis_c(option = "mako") + 
  theme_classic() + theme(legend.position = "none")

p | q
```

### Localisation in context

While pairwise proximity statistics provide a sensible first step, they can sometimes be misleading because they ignore the broader tissue context. For example, the K-cross function from `spicyR` indicates whether two cell types are found close together more often than expected under a random distribution, but it does not address conditional questions. Biologists, however, are often interested in questions such as:

*“Are there certain immune cells that cluster more tightly with tumor cells, after accounting for the fact that immune cells generally do not cluster with tumors?”*

This is where `Kontextual`, from the `r BiocStyle::Biocpkg("Statial")` package, becomes useful. `Kontextual` goes beyond pairwise enrichment by constructing a contextual null model. Rather than comparing two cell types against a fully random distribution, it compares the observed association to what would be expected given the local mixture of neighboring cell types. In practice, this means that `Kontextual` tests whether a particular association is stronger or weaker than expected given the surrounding cellular environment.

For this demonstration, we define an immune context composed of T cells, myeloid cells, and B cells. `Kontextual` then evaluates whether, within this immune backdrop, specific pairs (for example, myeloid–tumor) show evidence of additional attraction or repulsion at a chosen spatial scale. The method is flexible: any grouping of interest can be defined as a “parent” or context, allowing conditional relationships of a population to be interrogated relative to that context.

```{r ktx-run}
## Create a vector saying which cells are in the immune context
immune <- c("T_cells", "Myeloid", "B_cells")

## Define the parents we want to use as contexts
parentDf <- parentCombinations(
  all = unique(vhdsub$cellType),
  immune = immune)

## Run Kontextual with radius of 20
resultsKontextual <- Kontextual(
  cells = vhdsub,
  parentDf = parentDf,
  imageID = "sample_id",
  cellType = "cellType",
  spatialCoords = c("pxl_col_in_fullres", "pxl_row_in_fullres"),
  r = 200
)
```

Let's visualize the results.

```{r ktx-scatter-plot, fig.height=5, fig.width=8}
plotKontext <- resultsKontextual |>
  ggplot(aes(original, kontextual, label = test)) +
  geom_point() +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) +
  theme_classic()

ggplotly(plotKontext)
```

We can then inspect radius-dependent curves (for example, Myeloid cells around Tumor cells) to determine the spatial scales at which the effect emerges, which is often informative for processes such as invasion fronts or peritumoral niches.

```{r ktx-scale-curve, fig.height=4.6, fig.width=4.5, warning=FALSE, message=FALSE}
curves <- kontextCurve(
  cells = vhdsub, rs = seq(20, 500, 50), 
  from = "Tumor", to = "Myeloid", parent = immune, se = TRUE,
  image = "sample01", cellType = "cellType", imageID = "sample_id"
)

p <- kontextPlot(curves) + theme_classic() + 
  scale_color_manual(labels = c("Kontextual", "L-Function"), 
                     values = c("#00BE70", "darkblue")) + 
  scale_fill_manual(values = c("lightgreen", "lightblue")) + 
  theme(panel.grid.major = element_line(colour = "grey90", linetype = 2),
        legend.position = "top", 
        legend.text = element_text(size = 13), 
        axis.title = element_text(size = 14)) 
p@layers$geom_hline$aes_params$colour <- "black" # update dashed line color
p$layers[["geom_ribbon"]]$show.legend <- FALSE   # remove geom ribbon box
p

```

At a global level, the negative L-function values suggest that Myeloid and Tumor cells are spatially segregated. However, the positive `Kontextual` values indicate that Myeloid cells cluster more strongly with Tumor cells than would be expected given their immune context (i.e., relative to other immune cells).

To aid interpretation, we can also re-plot the spatial map with the immune context shown in gray and the two focal cell types highlighted in color. This helps link the statistical findings to tangible spatial patterns within the tissue.

```{r ktx-spatial-highlight, fig.height=3, fig.width=4, message=FALSE}
ct_interest <- c("Myeloid", "Tumor")
vhdsub$myetu <- ifelse(!(vhdsub$cellType %in% ct_interest), 
                       "Other", vhdsub$cellType)
vhdsub$myetu <- factor(vhdsub$myetu, levels = c(ct_interest, "Other"))
plotSpatialFeature(vhdsub, features = "myetu", 
                   colGeometryName = "updatecellSeg") + 
  scale_fill_manual(name   = "cellType",
                    values = c("Myeloid" = "#f58231", "Tumor" = "#bfef45",
                               "Other" = "grey90")) + 
  thin_discrete_legend
```

```{r hidden8, echo = FALSE, results = "hide"}
gc()
```

## Identifying expression changes associated with cell localisation

Spatial proximity tells us where different cell types are relative to one another, but an equally important question is what happens to those cells when they are in close contact. Proximity alone does not guarantee functional interaction. To uncover biological relevance, we need to test whether changes in the microenvironment are associated with measurable shifts in gene expression within a given cell type.

This is the motivation behind `spatioMark`, part of the `Statial` package. The idea is simple:

-   take a target population (for example, Myeloid cells),

-   quantify its local environment (how many Tumor cells, T cells, Fibroblasts, etc. are nearby within a certain radius),

-   and ask whether the transcriptomes of the target cells vary with that environment.

In practice, `spatioMark` first computes local abundances of each cell type around every cell, giving each cell a numerical “neighborhood profile.” These covariates then feed into a gene-wise regression model. For each gene expressed in the target population, `spatioMark` fits a generalized linear model. As we have count data in this case, we leverage functionality in the `edgeR` package to model the data as negative binomial. The coefficients tell us whether gene expression increases or decreases as the abundance of a chosen neighbor rises.

### Abundance

First, we need to quantify the abundance of each cell type around each cell.

```{r smark-abundances}
vhdsub <- getAbundances(vhdsub,
  r = 200,
  cellType = "cellType",
  imageID = "sample_id"
)
```

Next, we focus on whether Myeloid cells show distinct expression patterns when they are located close to Tumor cells.

```{r smark-statechanges-fit}
stateChanges <- calcStateChanges(
  cells = vhdsub,
  type = "abundances",
  from = "Myeloid",
  to = "Tumor",
  test = "nb",
  imageID = "sample_id"
)

stateChanges |> head(20)
```

The previous output is a ranked list of genes, highlighting markers that change as Myeloid cells are near Tumor cells. One of the top genes on this list is *SPP1* which is known to be expressed in tumor-associated macrophages. A positive coefficient of *SPP1* suggests higher expression of *SPP1* in Myeloid cells when they are in areas of high tumour density. We can visualize this relationship using the `plotStateChanges()` function.

```{r smark-plot-spp1, warning=FALSE, message=FALSE, fig.height=4,fig.width=4.7}
p <- plotStateChanges(
  cells = vhdsub,
  image = "sample01",
  type = "abundances",
  from = "Myeloid",
  to = "Tumor",
  marker = "SPP1",
  imageID = "sample_id",
  size = 1,
  shape = 19,
  interactive = FALSE,
  method = "lm"
)

p[[1]] + theme_void() + 
  theme(legend.key.width = grid::unit(0.5, "lines"), 
        legend.key.height = grid::unit(1, "lines")) + facet_null() + 
ggtitle("Cell Points: Myeloid, Cell Density: Tumor") + center_title 
```

### Contamination

However, some genes should not be expressed in macrophages, such as *CEACAM6*, and are therefore likely the result of lateral spillover or transcript diffusion.

```{r smark-ceacam6-check, fig.height=4, fig.width=3.5}
sce[, down_sample] |>
  join_features(features = "CEACAM6", shape = "wide") |>
  mutate(logCEACAM6 = log1p(CEACAM6)) |>
  ggplot(aes(x = Level1, y = logCEACAM6, fill = Level1)) +
  geom_boxplot() + 
  scale_fill_manual(values = unname(pals::trubetskoy())) +
  theme_classic() + ylab("CEACAM6 log expression") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  thin_discrete_legend
```

Because we already computed `RCTD` weights, we can mitigate contamination by adding deconvolution-derived components as covariates to our models. By storing the `RCTD` cell-type weight matrix into `reducedDim(vhdsub, "RCTD")`, we can pass `"RCTD"` to the contamination argument of `calcStateChanges()`. `spatioMark` will use these estimates of contamination in its modelling to hopefully remove false positives like *CEACAM6* while preserving genuine relationships.

```{r smark-add-contam-rctd}
contam <- assay(rctd_result_level1, 1) |>
  t() |>
  as.matrix() |>
  as.data.frame()

reducedDim(vhdsub, "RCTD") <- contam[match(colnames(vhdsub), rownames(contam)),]
```

After correction, the false discovery rate of *CEACAM6* is no longer significant at a threshold of 0.05. The rank of this gene falls from 14 to 173, while other macrophage-relevant top genes, such as *SPP1*, keep their high rankings and remain contamination-invariant.

```{r smark-statechanges-contam}
stateChangesCon <- calcStateChanges(
  cells = vhdsub,
  type = "abundances",
  from = "Myeloid",
  to = "Tumor",
  test = "nb",
  imageID = "sample_id",
  contamination = "RCTD"
)

stateChangesCon |> filter(marker == "CEACAM6")
```

Next, we can visualize the change of ranking of *CEACAM6* after decontamination, among the top 20 marker genes as Myeloid cells get near Tumor cells.

```{r rnkplot, fig.height=4, fig.width=5.2}
plotRankChange(target_gene = "CEACAM6", n = 20, stateChanges, stateChangesCon)
```

So far, we have analyzed one CRC sample from patient P2. In fact, @oliveira2025high includes multiple patients in this Visium HD public dataset, spanning different disease conditions (healthy vs. cancer). To apply spatial statistics to assess cell-type colocalization across multiple conditions, you can refer to the *Multi-sample* chapters of the OSTA book: a. [Differential spatial patterns](https://bioconductor.org/books/release/OSTA/pages/mult-diff-spatial-patterns.html), b. [Differential colocalization](https://bioconductor.org/books/release/OSTA/pages/mult-diff-colocalization.html), c. [Structure-based analysis](https://bioconductor.org/books/release/OSTA/pages/mult-structure-based.html)

```{r hidden9, echo = FALSE, results = "hide"}
gc()
```

## Extension

You can try using a finer resolution of cell type classified by `RCTD`.

```{r level2_annotation, eval=FALSE}
load(here::here("data/rctd_result_level2.RData"))
```

Or you can try using a different deconvolution algorithm.

```{r CARD, eval=FALSE}
set.seed(1000)
vhd2 <- vhdsub
colnames(spatialCoords(vhd2)) <- c("x", "y")
sce2 <-sce[, down_sample]
counts(vhd2) <- as(counts(vhd2), "sparseMatrix")
counts(sce2) <- as(counts(sce2), "sparseMatrix")
CARD_obj <- CARD_deconvolution(
    spe = vhd2,
    sce = sce2,
    sc_count = NULL,
    sc_meta = NULL,
    spatial_count = NULL,
    spatial_location = NULL,
    ct_varname = "Level2",
    ct_select = NULL,       # decon with all sce$Annogrp cell types
    sample_varname = NULL,  # use all sce as one ref sample
    mincountgene = 0,
    mincountspot = 0
)

# a matrix/data.frame of probabilities per cell (rows) × cell types (cols)
P <- CARD_obj$Proportion_CARD[colnames(vhdsub), ]

# cell type with the largest probability in each row
vhdsub$cellType2 <- colnames(P)[max.col(P, ties.method = "first")]

reducedDim(vhdsub, "CARD") <- as.data.frame(P)
```
